---
source: Rmd
title: "1. Finding your taxa in the Open Tree of Life Taxonomy"
teaching: 5
exercises: 5
questions:
- "What is the Open Tree Taxonomy?"
- "What are OTT ids?"
- 'What does TNRS stand for?'
objectives:
- "Getting OTT ids for some taxa."
- "Understanding TNRS approximate matching and its sensitivity."
# - "Finding the utility of taxonomic contexts"
# - 'Discovering functions to handle a "match_names" object.'
keypoints:
- "OTT ids are the Open Tree of Life Taxonomy handlers, they allow us to interact with the Open Tree synthetic tree."
# - "taxonomic context is very important to find the correct _ott ids_ for our taxa."
- "You can go from a scientific name to an OTT id with TNRS matching."
- "You cannot go from a common name to OTT id with the Open Tree of Life tools."
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
```
<br/>
<br/>

<!-- ### The Open Tree Taxonomy and its identifiers -->

The [Open Tree Taxonomy](https://tree.opentreeoflife.org/about/taxonomy-version/ott3.2) (or **OTT**) synthesizes taxonomic information and assigns each taxon a unique identifier, which we know as the **OTT id**.

To deal with synonyms and scientific name misspellings, the Open Tree Taxonomy uses
the [Taxonomic Name Resolution Service](http://tnrs.iplantcollaborative.org/) (**TNRS**
from now on), that allows linking scientific names to a unique OTT id when necessary.

To interact with the OTT, we will learn how to use a number of function from the `rotl` package.
Go to setup if you need to check if you have `rotl` installed or if you need to install it.

To get OTT ids for a set of taxa we will use the `rotl` function `tnrs_match_names()`.

This function will get you a **match_names object**, which is a data frame with a certain number of columns. Let's generate a match_names object and check it out.


```{r, eval = TRUE}
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
class(resolved_names)
resolved_names
```
<br/>
<br/>

> ## **Challenge!** Get the OTT ids from a match_names object
>
> You now have a match_names object that we called resolved_names. There are two main ways to extract the OTT ids from it. Can you figure them out?
>
> > ## Solution
> >
> > As a list, with the function `ott_id()`:
> >
> > ```{r}
> > rotl::ott_id(resolved_names)
> > ```
> >
> > Or, as a vector:
> >
> > ```{r}
> > resolved_names$ott_id
> > ```
> >
> > Most of the functions we will explore here take an argument called _ott ids_ as a numeric vector, so I will usually prefer the second approach.
> {: .solution}
{: .challenge}

<br/>
<br/>

> ## **Hack!**
>
> To facilitate the use of OTT ids later, you can name the rows of your match_names object.
>
> ```{r}
> rownames(resolved_names) <- resolved_names$unique_name
> ```
>
> This will allow an easy access to elements of the match_names object by calling the taxon name.
>
> ```{r, results = "asis"}
> resolved_names["Amphibia",]$ott_id
> # or
> resolved_names["Amphibia","ott_id"]
> ```
{: .callout}

<br/>
<br/>

> ## **Extra!** Accessing other elements of a match_names object
>
> Most elements of a match_names object can only be accessed as a vector, using the "$" symbol.
>
> The only other element that can be accessed as a list are **flags**.
>
> ```{r}
> rotl::flags(resolved_names)
> ```
{: .testimonial}

<!-- ## **Challenge!** What do other elements of a "match_names" object tell us?

## Solution
{: .solution}
{: .challenge}


### Taxonomic contexts and updating a match_names object

Running a tnrs match can take some time, so if you just want to update a few elements from your match_names object, you can use the functions `inspect()` and `update()`.

The _Mus_ example is fixed.

## **Hack!**

Put together two "match_names" objects with `c()` or `rbind()`
{: .callout} -->
<br/>
<br/>

> ## **Extra!**
>
> Going from common name to scientific name can be achieved with other tools. Check out [phylotastic](https://phylo.cs.nmsu.edu/) or [OneZoom](https://github.com/OneZoom/OZtree) APIs for that.
{: .testimonial}

<br/>
<br/>

{% include links.md %}
