---
source: Rmd
title: "Finding your taxa in the Open Tree of Life Taxonomy"
teaching: 5
exercises: 5
questions:
- "What is the Open Tree Taxonomy?"
- "What are _ott ids_?"
- "How do I get them?"
- "How do I check them and fix them?"
- "How do I use them?"
objectives:
- 'Learn about "tnrs_" functions, their arguments and their outputs'
- "Learn about approximate matching ans its sensitivity "
- "Learn about taxonomic contexts and how to use them"
keypoints:
- "_ott ids_ are the Open Tree of Life Taxonomy handlers, they allow us to interact with the Open Tree synthetic tree"
- "taxonomic context is very important to find the correct _ott ids_ of our taxa"
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
```

The Open Tree Taxonomy is


To get the _ott ids_ of your taxa you will need the function `tnrs_match_names()`. This will generate a "match_names" object.
```{r, eval = TRUE}
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
class(resolved_names)
resolved_names
```

Are you familiar with the elements of a match_names object?


There are two ways to extract the _ott ids_ from a match_names object.

As a list, with the function `ott_id()`:

```{r}
rotl::ott_id(resolved_names) # as a list
```

Or, as a vector:

```{r}
resolved_names$ott_id # as a vector
```

Most of the functions we will use need _ott ids_ as a vector, so I will prefer the second approach.

{% include links.md %}
