---
source: Rmd
title: "Finding your taxa in the Open Tree of Life Taxonomy"
teaching: 5
exercises: 5
questions:
- "What is the Open Tree Taxonomy?"
- "What are _ott ids_?"
- 'What is "tnrs"?'
objectives:
- "Getting ott ids for some taxa."
- "Understanding tnrs approximate matching and its sensitivity."
- "Finding the utility of taxonomic contexts"
- 'Discovering functions to handle a "match_names" object.'
keypoints:
- "_ott ids_ are the Open Tree of Life Taxonomy handlers, they allow us to interact with the Open Tree synthetic tree."
- "taxonomic context is very important to find the correct _ott ids_ for our taxa."
- "You can go from a scientific name to an _ott id_ with tnrs matching."
- "But you can go from common name to scientific name format using other functions, to make your own workflow."
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
```

### The Open Tree Taxonomy and its identifiers

The Open Tree Taxonomy stores information about taxa and assigns eaxh taxon a unique identifies which we cal the **_ott id_**.


To get the _ott ids_ of your taxa we will use the function `tnrs_match_names()`. This function generates a "match_names" object.

"tnrs" stands for Taxonomic Name Resolution Service.

```{r, eval = TRUE}
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
class(resolved_names)
resolved_names
```

> ## **Challenge!** Get the _ott ids_ from a "match_names" object
>
> There are two main ways to extract the _ott ids_ from a "match_names" object. Can you figure them out?
>
> > ## Solution
> >
> > As a list, with the function `ott_id()`:
> >
> > ```{r}
> > rotl::ott_id(resolved_names)
> > ```
> >
> > Or, as a vector:
> >
> > ```{r}
> > resolved_names$ott_id
> > ```
> >
> > Most of the functions we will explore take an argument of _ott ids_ as a numeric vector, so I will usually prefer the second approach.
> {: .solution}
{: .challenge}


> ## **Hack!**
>
> To facilitate the use of _ott ids_ later, you can name the rows of your "match_names" object.
>
> ```{r}
> rownames(resolved_names) <- resolved_names$unique_name
> ```
>
> This will allow an easy access to elements of the "match_names" object by calling the taxon name.
>
> ```{r}
> resolved_names["Amphibia",]$ott_id # or resolved_names["Amphibia","ott_id"]
> ```
{: .callout}

> ## **Extra!** Accessing other elements of a "match_names" object
>
> Most elements of a "match_names" object can only be accessed as a vector, using the "$" symbol.
>
> The only other element that can be accessed as a list are "flags"
>
> ```{r}
> rotl::flags(resolved_names)
> ```
{: .testimonial}

<!-- ## **Challenge!** What do other elements of a "match_names" object tell us?

## Solution
{: .solution}
{: .challenge}


### Taxonomic contexts and updating a "match_names" object

Running a tnrs match can take some time, so if you just want to update a few elements from your match_names object, you can use the functions `inspect()` and `update()`.

The _Mus_ example is fixed.

## **Hack!**

Put together two "match_names" objects with `c()` or `rbind()`
{: .callout} -->

> ## **Extra!**
>
> Going from common name to scientific name can be achieved with other tools. Check out [phylotastic](https://phylo.cs.nmsu.edu/) or [OneZoom](https://github.com/OneZoom/OZtree) APIs for that.
{: .testimonial}


{% include links.md %}
