---
source: Rmd
title: "Getting a piece of the Synthetic Open Tree of Life"
teaching: 5
exercises: 5
questions:
- "What is the synthetic Open Tree of Life?"
- "How do I interact with it?"
- "Why is my taxon not in the tree?"
objectives:
- "Extract an induced subtree"
- "Extract a subtree"
keypoints:
- "dropping singletons warning is ..."
- ""
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
```

Now that we have some _ott ids_ we can interact with the synthetic Open Tree.
For that, we can use two different functions that start with "tol_".

Let's get a tree of taxa from different ranks.

```{r, eval = TRUE, results = 'hide'}
mytree <- rotl::tol_induced_subtree(resolved_names$ott_id)
```

What does this warning mean?


```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE}
ape::plot.phylo(mytree, cex = 1)
```

This is cool!

Now, what if you want a synthetic tree containing some or all descendants from your taxa of interest?

# Get a subtree of one taxon

We can extract a subtree of all descendants of one taxon at a time using the function `tol_subtree()` and the amphibians _ott id_.

Try to extract a subtree of all amphibians.
Get the _ott id_ first. It is already stored in the `resolved_names` object, but you can run the function `tnrs_match_names()` again if you want.
```{r, eval = TRUE}
amphibia_ott_id <- resolved_names$ott_id[1] # extract the ott id from resolved_names
# amphibia_ott_id <- rotl::tnrs_match_names("amphibians")$ott_id # OR run tnrs again
```

Now, extract the subtree from the Open Tree synthetic tree using `tol_subtree()`.
```{r, eval = TRUE, results= 'hide', warning= FALSE}
amphibia_subtree <- rotl::tol_subtree(ott_id = amphibia_ott_id)
```
```{r, eval = TRUE}
amphibia_subtree
```
This is a large tree!

Try to extract a subtree for the genus _Canis_. It should be way smaller!

```{r, eval = TRUE}
canis <- rotl::tnrs_match_names("canis")  # get the ott id of the genus Canis (dogs)
canis
```
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
subtree <- rotl::tol_subtree(canis$ott_id) # get the subtree
```

Why am I getting an error saying that the _node id_ was not found and that my taxon is "broken"??

Generally, this happens when phylogenetic information does not match taxonomic information.
For example, extinct lineages are sometimes phylogenetically included within the taxon but are taxonomically excluded, making the taxon appear as paraphyletic.
On the browser, you can still get to the subtree. From R, you need to do something else first.
