---
source: Rmd
title: "Getting a piece of the Synthetic Open Tree of Life"
teaching: 5
exercises: 5
questions:
- "What is the synthetic Open Tree of Life?"
- "How do I interact with it?"
- "Why is my taxon not in the tree?"
objectives:
- "Get an induced subtree"
- "Get a subtree"
keypoints:
- "OTT ids and node ids allow us to interact with the synthetic OpenTree."
- "Portions of the synthetic OpenTree can be extracted from a single OTT id or from a bunch of OTT ids"
- "It is not possible to get a subtree from an OTT id that is not in the synthetic tree."
---
```{r, eval = TRUE, echo = FALSE, results = 'hide', warning = FALSE}
source("../bin/chunk-options.R")
tol <- rotl::tol_about()
```
<br/>
<br/>

The synthetic Open Tree of Life (**synthetic OpenTree** from now on) summarizes information from **`r tol$num_source_trees`**
trees from **`r tol$num_source_studies`** peer-reviewed and published studies, that have been uploaded to the OpenTree database (the Phylesystem) through a [curator system](https://tree.opentreeoflife.org/curator).

Functions from the `rotl` package that interact with the synthetic OpenTree start with `tol_`.

To access general information about the current synthetic OpenTree, we can use the function `tol_about()`. This function requires no argument.

```{r, eval = TRUE}
rotl::tol_about()
```
This is nice!

As you can note, the current synthetic OpenTree was created not too long ago, on `r tol$date_created`.

This is also telling us that there are currently more than **2 million tips** on the synthetic OpenTree.

It is indeed a large tree. So, **_what if we just want a small piece of the whole synthetic OpenTree?_**

Well, now that we have some interesting taxon OTT ids, we can easily do this.


### Getting an induced subtree

The function `tol_induced_subtree()` allows us to get a tree of taxa from different taxonomic ranks.

```{r, eval = TRUE, results = 'hide'}
my_tree <- rotl::tol_induced_subtree(resolved_names$ott_id)
```

<br/>

> ## Note: What does this warning mean?
>
> This warning has to do with the way the synthetic OpenTree is generated.
> You can look at the [overview of the synthesis algorithm](https://docs.google.com/presentation/d/1RwoNTUK3LKgBupBNOc1TNsIpucuMUddlYW6rC56We10/edit?usp=sharing) for more information.
>
>
{: .discussion}

<br/>

Let's look at the output of `tol_induced_subtree()`.

```{r, eval = TRUE}
my_tree
```

R is telling us that we have a rooted tree with no branch lengths and `r ape::Ntip(my_tree)` tips.
If we check the class of the output, we will verify that it is a **'phylo' object**.

```{r, eval = TRUE}
class(my_tree)
```

A 'phylo' object is a data structure that stores the necessary information to build a tree.
There are several functions from different packages to plot trees or 'phylo' objects in R (e.g., [phytools](https://cran.r-project.org/web/packages/phytools/phytools.pdf)). For now, we will use the one from the legendary `ape` package `plot.phylo()`:

```{r plot1, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE}
ape::plot.phylo(my_tree, cex = 2) # or just plot(my_tree, cex = 2)
```

This is cool!

But, why oh why did my _Canis_ disappear? `r emo::ji("cry")`

Well, it did not actually disappear, it was replaced by the label "`r my_tree$tip.label[2]`".

We will explain why this happens in the [next section](../03-broken-taxa/index.html).

Now, **_what if you want a piece of the synthetic OpenTree containing all descendants of your taxa of interest?_**

### Getting a subtree of one taxon

We can extract a subtree of all descendants of one taxon at a time using the function `tol_subtree()` and an OTT id of your choosing.

Let's extract a subtree of all amphibians.

First, get its OTT id. It is already stored in our `resolved_names` object:

```{r, eval = TRUE}
amphibia_ott_id <- resolved_names["Amphibia",]$ott_id
```
Or,  you can run the function `tnrs_match_names()` again if you want.
```{r, eval = FALSE}
amphibia_ott_id <- rotl::tnrs_match_names("amphibians")$ott_id
```

<br/>

Now, extract the subtree from the synthetic OpenTree using `tol_subtree()`.
```{r, eval = TRUE, results= 'hide', warning= FALSE}
amphibia_subtree <- rotl::tol_subtree(ott_id = resolved_names["Amphibia",]$ott_id)
```
Let's look at the output:
```{r, eval = TRUE}
amphibia_subtree
```
This is a large tree! We will have a hard time plotting it.

<br/>

Now, let's extract a subtree for the genus _Canis_. It should be way smaller!

```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
subtree <- rotl::tol_subtree(resolved_names["Canis",]$ott_id)
```

## `r paste(rep(emo::ji("scream"), 3), collapse = " ")`

What does this error mean??

A "broken" taxon error usually happens when phylogenetic information does not match taxonomic information.

For example, extinct lineages are sometimes phylogenetically included within a taxon but are taxonomically excluded, making the taxon appear as paraphyletic.

On the [Open Tree of Life browser](https://tree.opentreeoflife.org/opentree/argus/ottol@372706), we can still get to the subtree ([check it out here](https://tree.opentreeoflife.org/opentree/argus/opentree12.3@mrcaott47497ott110766)).

From R, we will need to do something else first. We will get to that on the next episode.

<br/>
