---
source: Rmd
title: "Getting a piece of the Synthetic Open Tree of Life"
teaching: 5
exercises: 5
questions:
- "What is the synthetic Open Tree of Life?"
- "How do I interact with it?"
- "Why is my taxon not in the tree?"
objectives:
- "Extract an induced subtree"
- "Extract a subtree"
keypoints:
- "dropping singletons warning is ..."
- ""
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
```

`tol_about()` needs no argument.

```{r, eval = TRUE}
tol <- rotl::tol_about()
```

Now that we have some interesting taxon _ott ids_ we can interact with the synthetic Open Tree.
For this purpose, we will use functions starting with "tol_".

### Get an induced subtree

The function `tol_induced_subtree() allows us to get a tree of various taxa at different taxonomic ranks.

```{r, eval = TRUE, results = 'hide'}
my_tree <- rotl::tol_induced_subtree(resolved_names$ott_id)
```

This is not an error, it is a warning. What does it mean?

```{r, eval = TRUE}
my_tree
```
```{r plot1, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE}
ape::plot.phylo(my_tree, cex = 2)
```

This is cool!

But, why did my _Canis_ disappeared?

Well, it did not actually disappear, it was replaced by the label "`r my_tree$tip.label[2]`".

We will explain why this happens in the next section.

Now, **_what if you want a synthetic tree containing some or all descendants from your taxa of interest?_**

### Get a subtree of one taxon

We can extract a subtree of all descendants of one taxon at a time using the function `tol_subtree()` and the amphibians _ott id_.

Try to extract a subtree of all amphibians.
Its _ott id_  is already stored in the `resolved_names` object, but you can run the function `tnrs_match_names()` again if you want.
```{r, eval = TRUE}
resolved_names["Amphibia",]$ott_id # extract the ott id from resolved_names
amphibia_ott_id <- rotl::tnrs_match_names("amphibians")$ott_id # OR run tnrs again
```

Now, extract the subtree from the Open Tree synthetic tree using `tol_subtree()`.
```{r, eval = TRUE, results= 'hide', warning= FALSE}
amphibia_subtree <- rotl::tol_subtree(ott_id = resolved_names["Amphibia",]$ott_id)
```
```{r, eval = TRUE}
amphibia_subtree
```
This is a large tree! We will have a hard time plotting it.

Try to extract a subtree for the genus _Canis_. It should be way smaller!

```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
subtree <- rotl::tol_subtree(resolved_names["Canis",]$ott_id)
```

## `r paste(rep(emo::ji("scream"), 3), collapse = " ")`

What does this error mean??

A "broken" taxon error usually happens when phylogenetic information does not match taxonomic information.

For example, extinct lineages are sometimes phylogenetically included within a taxon but are taxonomically excluded, making the taxon appear as paraphyletic.

On the [Open Tree of Life browser](https://tree.opentreeoflife.org/opentree/argus/ottol@372706), we can still get to the subtree ([check it out here](https://tree.opentreeoflife.org/opentree/argus/opentree12.3@mrcaott47497ott110766)).

From R, we will need to do something else first.
