---
source: Rmd
title: '3. Dealing with "broken" and "invalid" taxa'
teaching: 5
exercises: 5
questions:
- "How do I detect a broken taxon?"
objectives:
- "Use the function is_in_tree()"
- "Understand outputs from those functions"
keypoints:
- "It is not possible to get a subtre from an OTT id that is not in the synthetic tree."
- "OTT ids and node ids allow us to interact with the synthetic OTOL."
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
```

<br/>
<br/>

We say that a taxon is "broken" when its OTT id is not assigned to a node in the synthetic tree.
As mentioned before, this happens when the OTT id belongs to a taxon that is not monophyletic in the synthetic tree.
This is the reason why we get an error when we try to get a synthetic subtree including that OTT id: it is **not** in the tree.


There is a way to find out that a group is "broken" before trying to get the subtree and getting an error.
```{r, eval = TRUE, warning= FALSE, error = TRUE}
rotl::is_in_tree(resolved_names["Canis",]$ott_id)
```

Indeed, our _Canis_ is **not** in the synthetic OTOL. To extract a subtree of a "broken" taxon, we have some options. But we will focus on one.

### Getting the MRCA of a taxon

The function `tol_node_info()` gets for you all relevant information of the node that is the ancestor or MRCA of a taxon. That also includes the actual node id.
```{r, eval = TRUE, warning= FALSE, error = TRUE}
canis_node_info <- rotl::tol_node_info(resolved_names["Canis",]$ott_id)
canis_node_info
```
Let's explore the class of the output.

```{r, eval = TRUE, warning= FALSE, error = TRUE}
class(canis_node_info)
```

<br/>

So we have an object of class 'list' and 'tol_node'. When we printed it, we got some
information. But we do not know how much information might not be "printed" to screen.

<br/>

Let's use the functions `str()` or `ls()` to check out the data strcture of our 'tol_node' object.

```{r, eval = TRUE, warning= FALSE, error = TRUE}
str(canis_node_info)
```

This is telling us that `tol_node_info()` extracted 8 different pieces of information from my node.
Right now we are only interested in the node ir. Where do you think it is?

<br/>

> ## Hands on! Get the node id of _Canis_ MRCA
>
> Extract it from your `canis_node_info` object and call it `canis_node_id`.
>
> ```{r, eval = TRUE}
> canis_node_id <- canis_node_info$node_id
> ```
{: .challenge}

<!-- > > ## **Extra: tol_lineage()**
> >
> > `tol_lineage()` gets information from all ancestral nodes from a given node id.
> >
> > Setting up include_lineage = TRUE in `tol_node_info()` will call this function and include that information along the output that can be accessed with `tax_lineage()`.
> {: .solution}
{: .testimonial} -->
<br/>

> > ## Pro tip 3.1: Get the node id of the MRCA of a group of OTT ids
> >
> > Sometimes you want the MRCA of a bunch of lineages. The function `tol_mrca()` gets the node of the MRCA of a group of OTT ids.
> >
> > Can you use it to get the mrca of _Canis_?
> {: .solution}
{: .testimonial}


The _node_ that contains _Canis_ is "`r canis_node_info$node_id`".

<br/>

### Getting a subtree using a node id instead of the taxon OTT id

Now that we have a node id, we can use it to get a subtree with `tol_subtree()`, using the argument `node_id`.

```{r, eval = TRUE, results= 'hold', warning= FALSE, error = TRUE}
canis_node_subtree <- rotl::tol_subtree(node_id = canis_node_id)
canis_node_subtree
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 15}
ape::plot.phylo(canis_node_subtree, cex = 1.2)
```

Nice! We got a subtree of 85 tips, containing all descendants from the node that also contains _Canis_.

If you explore the taxon names at the tip, you will notice that this includes species assigned to genera other than _Canis_.

Now, what if I want a subtree of certain taxonomic ranks withing my group? Go to the next episode and find out how you can do this!


> > ## Pro Tip 3.2: Get an induced subtree of taxonomic children
> >
> > What if I _really, really_ need a tree containing species within the genus _Canis_ only, excluding everything that does not belong to the genus taxonomically, even if it does phylogenetically?
> >
> > We can get the OTT ids of the taxonomic children of our taxon of interest and use the function `tol_induced_subtree()`.
> >
> >
> > First, we will get the taxonomic children.
> > ```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
> > canis_taxonomy <- rotl::taxonomy_subtree(resolved_names["Canis",]$ott_id)
> > ```
> > ```{r, eval = TRUE, warning= FALSE, error = TRUE}
> > canis_taxonomy
> > ```
> > Now, extract the OTT ids.
> > ```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
> > canis_taxonomy_ott_ids <- datelife::extract_ott_ids(x = canis_taxonomy$tip_label)
> > ```
> > Try to get an induced subtree of _Canis_ taxonomic children.
> >
> > ```{r, eval = TRUE, results = 'hide', warning= FALSE, error = TRUE}
> > canis_taxonomy_subtree <- rotl::tol_induced_subtree(canis_taxonomy_ott_ids)
> > ```
> > It is often not possible to get an induced subtree of all taxonomic children from a taxon,
> > because some of them will not make it to the synthetic tree.
> >
> > To verify which ones are giving us trouble, we can use the function `is_in_tree()` again.
> > ```{r, eval = TRUE, warning= FALSE, error = TRUE, results = 'hide'}
> > canis_in_tree <- sapply(canis_taxonomy_ott_ids, rotl::is_in_tree) # logical vector
> > canis_taxonomy_ott_ids_intree <- canis_taxonomy_ott_ids[canis_in_tree] # extract ott ids in tree
> > ```
> >
> > Now get the tree.
> > ```{r, eval = TRUE, results = 'hide', warning= FALSE, error = TRUE}
> > canis_taxonomy_subtree <- rotl::tol_induced_subtree(canis_taxonomy_ott_ids_intree)
> > ```
> > Plot it.
> > ```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 9}
> > ape::plot.phylo(canis_taxonomy_subtree, cex = 1.2)
> > ```
> >
> > There! We have a synthetic subtree (derived from phylogenetic information) containing only the taxonomic children of _Canis_.
> {: .solution}
{: .testimonial}

<br/>
