---
source: Rmd
title: "4. Getting an induced subtree of all taxa within a taxonomic rank"
teaching: 5
exercises: 5
questions:
- "How do I get all taxa from a certain rank?"
objectives:
- "Get an induced subtree from all taxa of a certain taxonomic rank."
keypoints:
- "It is possible to get all types of subsets from the synthetic tree, as long as you can get the OTT ids!"
---
```{r, eval = TRUE, echo = FALSE, results = 'hide', warning = FALSE}
source("../bin/chunk-options.R")
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
amphibia_subtree <- rotl::tol_subtree(ott_id = resolved_names["Amphibia",]$ott_id)
```

<br/>
<br/>

There is not a specific function to get all taxa from a given rank.
Of course, we can always hack our way through it.
In the package `datelife`, we have implemented a function called `get_ott_children()`, that extracts OTT ids of all taxa from a rank specified by the argument `ott_rank`.

Let's get all amphibian families.

```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
amphibia_families <- datelife::get_ott_children(ott_ids = resolved_names["Amphibia",]$ott_id, ott_rank = "family")
```
Now, get the induced subtree using the amphibian families' OTT ids.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
amphibia_families_subtree <- rotl::tol_induced_subtree(amphibia_families$Amphibia$ott_id)
```

Look at the output.
```{r, eval = TRUE, warning= FALSE, error = TRUE, fig.height = 13}
amphibia_families_subtree
ape::plot.phylo(amphibia_families_subtree, cex = 1.2)
```
Super cool!

> ## Hands on! Get a family subtree without ott ids in the tip labels
>
> Hint: Look at the arguments of function `tol_induced_subtree()`
>
> > ## Solution
> > ```{r, eval = TRUE}
> > amphibia_families_subtree2 <- rotl::tol_induced_subtree(amphibia_families$Amphibia$ott_id, label_format = "name")
> > ape::plot.phylo(amphibia_families_subtree2, cex = 1.2)
> > ```
> {: .solution}
{: . challenge}

> ## Hack: Get all taxa from a taxonomic rank.
>
> There are several ways to do this using `rotl` functions only. Try it out!
>
> > ## Look at a possible solution
> >
> > ```{r, eval = TRUE}
> > amphibia_taxonomy <- rotl::taxonomy_subtree(resolved_names["Amphibia",]$ott_id)
> > ls(amphibia_taxonomy)
> > length(amphibia_taxonomy$tip_label)
> > head(amphibia_taxonomy$tip_label)
> > tail(amphibia_taxonomy$tip_label)
> > amphibia_taxonomy$edge_label
> > edges <- datelife::extract_ott_ids(x=amphibia_taxonomy$edge_label)
> > length(edges)
> > edges_taxon_info <- rotl::taxonomy_taxon_info(edges)
> > ls(edges_taxon_info[[1]])
> > is_family <- unname(unlist(sapply(edges_taxon_info, "[", "rank") %in% "family"))
> > is_suppressed <- unname(unlist(sapply(edges_taxon_info, "[", "is_suppressed_from_synth")))
> > # flag "is suppressed from synth" is not updated, so it is useless for now.
> > amphibia_families <- unname(unlist(sapply(edges_taxon_info, "[", "ott_id")[is_family]))
> > in_tree <- rotl::is_in_tree(amphibia_families)
> > amphibia_families_subtree <- rotl::tol_induced_subtree(amphibia_families[in_tree])
> > ```
> {: .solution}
{: .challenge}

We have seen up to now how to get a portion of the synthetic OTOL. How do I inspect the source phylogenetic trees that support the subtrees?

<br/>
