---
source: Rmd
title: "Getting branch length information (proportional to time) for you taxa"
teaching: 5
exercises: 5
questions:
- "How do I do this?"
objectives:
- "Understand the difference between find studies and datelife"
keypoints:
- "..."
---
```{r, eval = TRUE, echo = FALSE}
source("../bin/chunk-options.R")
```

You can explore the studies supporting the synthetic tree, and find out which ones have branch lengths.

In the package `datelife`, we have implemented a workflow that extracts all studies containing information from at least two taxa.
It only works at the species level.

```{r, eval = TRUE}
apes <- c("Pongo", "Pan", "Gorilla", "Hoolock", "Homo")
resolved_names <- rotl::tnrs_match_names(apes)
```

```{r, eval = TRUE, results = 'hide'}
apes_dr <- datelife::get_datelife_result(resolved_names$unique_name, get_spp_from_taxon = TRUE)
```
```{r, eval = TRUE}
names(apes_dr)
apes_dr[1] # look at the first element of the list
apes_dr[length(apes_dr)] # look at the last element of the list
```

We have now a list of matrices storing time of lineage divergence data for all taxon pairs.
Lists are named by the study citation, so we have that information handy at all times.
