---
source: Rmd
title: "Getting studies and trees supporting relationships in a synthetic subtree"
teaching: 5
exercises: 5
questions:
- "What are the original studies supporting relationships in my synthetic subtree?"
objectives:
- "Get supporting trees for certain regions of the synthetic Open Tree of Life."
keypoints:
- "Supporting trees usually contain more taxa than the ones we are interested in."
---
```{r, eval = TRUE, echo = FALSE, results = 'hide', warning = FALSE}
source("../bin/chunk-options.R")

```

<br/>
<br/>

To get the source trees supporting a node from our synthetic tree we will need two functions.
The function `source_list`() gets the study and tree ids (and other info) from source studies (not the trees). It is applied to a 'tol_node' object.

We already have one that we generated with `tol_node_info()`, do you remember how we called it?

> ## Hands on! Get all supporting trees.
> Get the supporting study metadata from the _Canis_ node info. Store it in an object called `canis_node_studies`.
> Look at its class and the information it contains.
>
> ```{r}
> canis_node_studies <- rotl::source_list(canis_node_info)
> ```
>
> ```{r}
> class(canis_node_studies)
> ```
>
> ```{r}
> str(canis_node_studies)
> ```
>
{: .challenge}

Now that we have the ids, we can use the function `get_study_tree()`, which will get us the actual supporting trees.
This function takes one _study id_ and _tree id_ at a time, like this:

```{r}
x <- 1
rotl::get_study_tree(study_id = canis_node_studies$study_id[x], tree_id = canis_node_studies$tree_id[x], tip_label="ott_taxon_name", deduplicate = TRUE)
```

> ## Hands on! Get all supporting trees.
>
> Call the output _canis_source_trees_
>
> Hint: You can use a **"for" loop** or an `apply()` function to get them all.
>
> > ## Solution
> >
> > With a 'for' loop.
> >
> > ```{r}
> > canis_source_trees <- vector(mode = "list") # generate an empty list
> > for (i in seq(nrow(canis_node_studies))){
> >   source_tree <- rotl::get_study_tree(study_id = canis_node_studies$study_id[i], tree_id = canis_node_studies$tree_id[i], tip_label="ott_taxon_name", deduplicate = TRUE)
> >   canis_source_trees <- c(canis_source_trees, list(source_tree))
> > }
> > canis_source_trees
> > ```
> >
> > With an `apply()` function.
> >
> > ```{r}
> > canis_source_trees <- sapply(seq(nrow(canis_node_studies)), function(i)
> >   rotl::get_study_tree(study_id = canis_node_studies$study_id[i], tree_id = canis_node_studies$tree_id[i], tip_label="ott_taxon_name", deduplicate = TRUE))
> > canis_source_trees
> > ```
> {: .solution}
{: .challenge}

The object `canis_node_studies` contains a lot of information. You can get it using a 'for' loop, or an `apply()` function.

A key piece of information are the citations from the supporting studies. We can get these for each source trees with the function `get_study_meta()`. Let's do it. First we need the study meta:

```{r, eval = TRUE, warning = FALSE}
canis_node_studies_meta <- lapply(seq(nrow(canis_node_studies)), function(i)
  rotl::get_study_meta(study_id = canis_node_studies$study_id[i]))
```

Now we can get the citations:

```{r, eval = TRUE, warning = FALSE}
canis_node_studies_citations <- sapply(seq(length(canis_node_studies_meta)), function (i) canis_node_studies_meta[[i]]$nexml$`^ot:studyPublicationReference`)
```

Finally, let's plot the supporting trees along with their citations.
```{r canis-support-trees, eval = TRUE, warning = FALSE, fig.height = 24}
for (i in seq(length(canis_source_trees))){
  print(paste("The supporting tree below has", length(canis_source_trees[[i]]$tip.label), "tips."))
  print(paste("Citation is:", canis_node_studies_citations[i]))
  ape::plot.phylo(canis_source_trees[[i]])
}
```

<br/>

Note that the supporting trees for a node can be larger than the subtree itself.

You will have to drop the unwanted taxa from the supporting studies if you just want the parts that belong to the subtree.

Moreover, the tip labels have different taxon names in the source trees and the synthetic subtrees. I you go to the browser, you can access original tips and matched tips, but R drops that info. We would have to standardize them with TNRS before trying to subset, and that takes some time and often visual inspection.

<br/>
