---
source: Rmd
title: "Getting studies and trees supporting relationships in a synthetic subtree"
teaching: 5
exercises: 5
questions:
- "What are the original studies supporting relationships in my synthetic subtree?"
objectives:
- "Get supporting trees for certain regions of the synthetic Open Tree of Life."
keypoints:
- "Supporting trees usually contains more taxa than the ones we are interested in."
---
```{r, eval = TRUE, echo = FALSE, results = 'hide', warning = FALSE}
source("../bin/chunk-options.R")
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
canis_node_info <- rotl::tol_node_info(resolved_names["Canis",]$ott_id)
canis_node_subtree <- rotl::tol_subtree(node_id = canis_node_info$node_id)
```

```{r}
canis_node_studies <- rotl::source_list(canis_node_info)
canis_node_studies
```
To get the actual trees, we will need the function `get_study_tree()`.
It takes one _study id_ and _tree id_ at a time.

```{r}
x <- 1
rotl::get_study_tree(study_id = canis_node_studies$study_id[x], tree_id = canis_node_studies$tree_id[x], tip_label="ott_taxon_name", deduplicate = TRUE)
```

> ## **Challenge!** Get all supporting trees.
>
> Call the output `canis_source_trees`
>
> Hint: You can use a "for" loop or an `apply()` function to get them all.
>
> > ## Solution
> >
> > With a "for" loop.
> >
> > ```{r}
> > canis_source_trees <- vector(mode = "list") # generate an empty list
> > for (i in seq(nrow(canis_node_studies))){
> >   source_tree <- rotl::get_study_tree(study_id = canis_node_studies$study_id[i], tree_id = canis_node_studies$tree_id[i], tip_label="ott_taxon_name", deduplicate = TRUE)
> >   canis_source_trees <- c(canis_source_trees, list(source_tree))
> > }
> > canis_source_trees
> > ```
> >
> > With an `apply()` function.
> >
> > ```{r}
> > canis_source_trees <- sapply(seq(nrow(canis_node_studies)), function(i)
> >   rotl::get_study_tree(study_id = canis_node_studies$study_id[i], tree_id = canis_node_studies$tree_id[i], tip_label="ott_taxon_name", deduplicate = TRUE))
> > canis_source_trees
> > ```
> {: .solution}
{: .challenge}

Get the citations for this source trees.
```{r, eval = TRUE, warning = FALSE}
canis_node_studies_meta <- lapply(seq(nrow(canis_node_studies)), function(i)
  rotl::get_study_meta(study_id = canis_node_studies$study_id[i]))
canis_node_studies_citations <- sapply(seq(length(canis_node_studies_meta)), function (i) canis_node_studies_meta[[i]]$nexml$`^ot:studyPublicationReference`)

```

Let's plot the supporting trees.
```{r, eval = TRUE, warning = FALSE, fig.height = 24}
for (i in seq(length(canis_source_trees))){
  ape::plot.phylo(canis_source_trees[[i]])
  print(paste("The chronogram above has", length(canis_source_trees[[i]]$tip.label), "tips."))
  print("Citations is:")
  print(canis_node_studies_citations[i])
}
```
Note that supporting trees for a node can be larger than the subtree itself.

You will have to drop the unwanted taxa from the supporting studies if you just want the parts that belong to the subtree.

Also, tip labels have different names, so we have to standardize them with tnrs before trying to subset.
