---
source: Rmd
title: "6. Getting branch length information (proportional to time) for you taxa"
teaching: 5
exercises: 5
questions:
- "How do I find supporting trees that include branch lengths?"
- "How do I subset them to include just the taxa I am interested in?"
objectives:
- "Learn about the `opentree_chronograms` object from datelife."
- "Get source chronograms from the `opentree_chronograms` object for a set of taxa."
keypoints:
- "datelife stores all chronograms from the Open Tree of Life phylesystem."
- "chronograms are stored in the `opentree_chronograms` object."
- "source chronograms are retrieved at the species level only (for now)."
---
```{r, eval = TRUE, echo = FALSE, warning = FALSE, results = 'hide'}
source("../bin/chunk-options.R")
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
canis_node_info <- rotl::tol_node_info(resolved_names["Canis",]$ott_id)
library(datelife)
all_properties <- rotl::studies_properties()
```

<br/>
<br/>

What if I want to search the OTOL database (phylesystem) for studies and trees matching some criteria?

The `rotl` package has functions that allow getting a list of studies or source trees matching specific criteria.

You will recognise these functions because they start with the word `studies_`.

Now, what kind of properties can we search for in the OTOL database?
The function `studies_properties()` gets for us two lists, one for studies and another one for tree properties available for search.

Take a look at them:
```{r, eval = TRUE}
rotl::studies_properties()
```

As you can see, the actual values that this properties can take are not available in the output of the function. Go to the [phylesystem API wiki](https://github.com/OpenTreeOfLife/phylesystem-api/wiki/NexSON) to get them, along with an explanation of their meaning.

<!-- `studies_find_studies()` -->


To get all trees with branch lengths poprotional to time we need the function
`studies_find_trees()`, using the property "ot:branchLengthMode" and the value "ot:time".
It takes some time for it to get all the information, so we will not do it now.
Go to [Instructor Notes](../guide/index.html###6.1) later for more information on how to do this.

<!--
> > ## Extra! Try it yourself.
> >
> > ```{r, eval = TRUE}
> > chronograms <- rotl::studies_find_trees(property = "ot:branchLengthMode", value = "ot:time", verbose = TRUE, detailed = TRUE)
> > ```
> > ```{r, eval = TRUE}
> > class(chronograms)
> > names(chronograms)
> > ```
> > We should be able to use `list_trees()` to get all trees matching our criteria.
> > ```{r, eval = TRUE}
> > rotl:::list_trees(chronograms)
> > ```
> >
> > Except, it does not really work.
> {: .solution}
{: .testimonial} -->


### Search the opentree chronogram database from `datelife`

In the package `datelife`, we have implemented a workflow that extracts all studies containing information from at least two taxa.

You can get all source chronograms from an induced subtree, as long as the tip labels
are in the "name" format (and not the default "name_and_id").

`datelife` takes as input either a tree with tip labels as scientific names (andd not names and ids), or a vector of scientific names.

Get a _Canis_ subtree with tip labels that do not contain the OTT id.

```{r, eval = TRUE, warning = FALSE, echo = FALSE}
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
canis_node_info <- rotl::tol_node_info(resolved_names["Canis",]$ott_id)
```

```{r, eval = TRUE, warning = FALSE}
canis_node_subtree <- rotl::tol_subtree(node_id = canis_node_info$node_id, label = "name")
canis_node_subtree
```

Now, you can use that tree as input for the `get_datelife_result()` function.

```{r, eval = TRUE}
canis_dr <- datelife::get_datelife_result(canis_node_subtree)
```
We have now a list of matrices storing time of lineage divergence data for all taxon pairs.

Lists are named by the study citation, so we have that information handy at all times.

Let's explore the output.

```{r, eval = TRUE}
names(canis_dr)
canis_dr[1] # look at the first element of the list
canis_dr[length(canis_dr)] # look at the last element of the list
```

### Get your chronograms

Then, it is really easy to go from a matrix to a tree, using the function `summarize_datelife_result()` with the option `summary_format = "phylo_all"`. Note the printed output returns a summary of taxa that have branch length information in the database.

```{r, eval = TRUE, warning = FALSE}
canis_phylo_all <-  datelife::summarize_datelife_result(canis_dr, summary_format = "phylo_all")
```

### Plot your results

To plot the resulting tree, you can use the `plot.phylo()` function from `ape`.
You can also use the function `plot_phylo_all()`, that adds the study citation as title and a geostratigraphic axis.

```{r, eval = TRUE, warning = FALSE, message = FALSE}
datelife::plot_phylo_all(trees = canis_phylo_all)
```

<br/>
<br/>
