---
source: Rmd
title: "Summarizing branch length information"
teaching: 5
exercises: 5
questions:
- "How do I summarize information from different source chronograms?"
- "How do I choose a preferred source chronogram?"
objectives:
- "Understanding the depth of uncertainty around age estimates."
keypoints:
- "Source chronograms have a wide range of variation in age estimates."
---
```{r, eval = TRUE, echo = FALSE, message = FALSE, results = 'hide', warning = FALSE}
source("../bin/chunk-options.R")
# apes <- c("Pongo", "Pan", "Gorilla", "Hoolock", "Homo")
# resolved_names <- rotl::tnrs_match_names(apes)
# apes_dr <- datelife::get_datelife_result(resolved_names$unique_name, get_spp_from_taxon = TRUE)
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
rownames(resolved_names) <- resolved_names$unique_name
canis_node_info <- rotl::tol_node_info(resolved_names["Canis",]$ott_id)
canis_node_subtree <- rotl::tol_subtree(node_id = canis_node_info$node_id, label = "name")
canis_dr <- datelife::get_datelife_result(canis_node_subtree)
```

<br/>
<br/>

Now that we have a collection of chronograms containing our taxa of interest, we can go on to summarize the information in them.

There is no consensus on the best way to do this.

We have implemented two ways of summarizing information from several chronograms into a single one.
The fastest one is using the median of node ages for each node with available information, and then evenly distributing ages across nodes.

```{r, eval = TRUE, results = 'hide', message = FALSE}
canis_phylo_median <-  datelife::summarize_datelife_result(canis_dr, summary_format = "phylo_median")
```

<br/>
<br/>

Check that we actually went from a list of matrices to a tree with branch lengths:
```{r eval = TRUE}
canis_phylo_median
````

<br/>
<br/>

Good. Now we can plot our chronogram!

```{r plot60, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 10}
ape::plot.phylo(canis_phylo_median, cex = 1.2)
# Add the time axis:
ape::axisPhylo()
# And a little hack to add the axis name:
graphics::mtext("Time (myrs)", side = 1, line = 2, at = max(get("last_plot.phylo",envir = .PlotPhyloEnv)$xx) * 0.5)
````

> ## Challenge! Get the other type of summary chronogram
>
> Hint: Explore options from the argument **summary_format** in the function `summarize_datelife_result()`
>
> >
> > ## Solution
> > ```{r, eval = TRUE, results = 'hide', message = FALSE}
> > canis_phylo_sdm <-  datelife::summarize_datelife_result(canis_dr, summary_format = "phylo_sdm")
> > ```
> > ```{r, eval = TRUE}
> > canis_phylo_sdm
> > ```
> > ```{r plot61, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 10}
> > ape::plot.phylo(canis_phylo_sdm, cex = 1.2)
> > ape::axisPhylo()
> > graphics::mtext("Time (myrs)", side = 1, line = 2, at = max(get("last_plot.phylo",envir = .PlotPhyloEnv)$xx) * 0.5)
> > ````
> >
> > As you can note, the SDM sumary chronogram is slightly older than the median summary chronogram!
> {: .solution}
{: .challenge}

<br/>
<br/>

Finally, give it a try on the web browser of [datelife](http://datelife.org/query/), too.
You can do the same things using a graphical user interface. It is fun!

<br/>
<br/>
